---
title: "ESM 244 lab 2"
author: "Amelia Ritger"
date: "1/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(janitor)
library(naniar)
library(VIM)
library(skimr)
library(ggfortify)
```

## 1. Get some data
```{r}
ca_pb <- read_csv(here("data", "ca_pollution_burden.csv"))
ca_dem <- read_csv(here("data", "ca_census_demographics_2010.csv"))
```

## 2. Some data cleaning and wrangling
```{r}
ca_pb_nopct <- ca_pb %>% 
  clean_names() %>%
  select(-contains("perc")) %>% 
  select(-contains("pctl")) %>%
  select(-latitude, -longitude)
```

```{r}
ca_dem_clean <- ca_dem %>% 
  clean_names()
```

## 3. Pollution burden indicators only 
(create PCA specific selected dataset)
```{r}
ca_pb_subset <- ca_pb_nopct %>% 
  select(ozone:solid_waste, asthma:housing_burden)
```

Now let's do PCA
Use the `prcomp()` function
```{r, eval=FALSE}
pb_PCA <- prcomp(ca_pb_subset, scale=TRUE)
```
This gives an error because there are missing values!

Aside: Explore missingness
```{r}
summary(ca_pb_subset) #there are 242 NAs in the data
#but that's clunky, here's a nicer way to visualize NAs:
naniar::gg_miss_var(ca_pb_subset)
matrixplot(ca_pb_subset, sortby="poverty") #shade indicates value level; red indicates missing value (NA)
```

Remove by complete case deletion (any observation with any NA values)
```{r}
ca_pb_nona <- ca_pb_subset %>% 
  drop_na()
skimr::skim(ca_pb_nona) #alternative to summary()
```

No NA values now. Cool. Ready to do PCA (otherwise, consider imputation, etc)
```{r}
my_ca_pca <- prcomp(ca_pb_nona, scale=TRUE)
summary(my_ca_pca)

#biplot(my_ca_pca)
#looks like a car wreck, let's try autoplot()

my_biplot <- autoplot(my_ca_pca,
                      colour = NA,
                      loadings.label = TRUE,
                      loadings.label.size = 3,
                      loadings.label.colour = "black",
                      loadings.label.repel = TRUE)
```

## 4. PCA for pollution burden + demographics

Join together the ca_dem_clean with ca_pb_nopct

```{r}
ca_df <- ca_dem_clean %>% 
  inner_join(ca_pb_nopct, by=c("census_tract_number" = "census_tract"))
```

##### At this point I left lab

Check it out, then get complete cases:
```{r}
ca_df_nona <- ca_df %>% 
  drop_na()
```

## 5. Make a new subset for PCA, that includes % white and elderly, and some interesting pollution burden & health indicators:

Like (you can choose a different set): 

- white_percent
- elderly_65_percent
- pm2_5
- pesticides
- traffic
- asthma
- cardiovascular_disease
- poverty

Make our subset:
```{r}
my_sub <- ca_df_nona %>% 
  select(white_percent, elderly_65_percent, pm2_5, pesticides, traffic, asthma, cardiovascular_disease, poverty)
```

Then run PCA: 
```{r}
my_dem_pca <- prcomp(my_sub, scale = TRUE)
biplot(my_dem_pca) # Noooo. We need better
```

Check it out a bit: 
```{r}
# Proportion of variance (& cumulative variance) explained by each PC
summary(my_dem_pca)

# Rotations (linear combinations for each PC):
my_dem_pca
```

Make a sweet biplot:
```{r}
my_dem_biplot <- autoplot(my_dem_pca, 
                      colour = NA,
                      loadings.label = TRUE,
                      loadings.label.size = 3,
                      loadings.label.colour = "black",
                      loadings.label.repel = TRUE) +
  theme_minimal() +
  scale_y_continuous(limits = c(-0.05, 0.05))

my_dem_biplot
```

